\documentclass[10pt, a4paper]{article}
\usepackage{lrec2016}
\usepackage{multibib}
\newcites{languageresource}{Language Resources}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage{apacite}
\usepackage{covington}
\usepackage{tikz-qtree}
\usepackage{tikz-dependency}
\usepackage{ctable}
\usepackage{url}

\usepackage{relsize}
%\bibliographystyle{apacite}

\let\w=\emph
\let\tag=\textsf
\let\app=\textsc
\newcommand\deprel[1]{\\\textsc{#1}}

\tikzset{edge from parent/.style={->,draw,font={\scshape\small}},
    every tree node/.style={align=center,anchor=north},
    level distance=10ex}

\setlength\textfloatsep{3.5mm}

%\title{Norwegian Universal Dependencies}
\title{Universal Dependencies for Norwegian}

\name{Lilja {\O}vrelid,\hspace{0.1cm} Petter Hohle}

\address{Department of Informatics,\\ University of Oslo \\
\texttt{$\{$liljao,pettehoh$\}$@ifi.uio.no}}

\date{}

\frenchspacing

\abstract{ This article describes the conversion of the Norwegian
  Dependency Treebank to the Universal Dependencies scheme. This paper
  details the mapping of PoS tags, morphological features and
  dependency relations and provides a description of the structural
  changes made to NDT analyses in order to make it compliant with the
  UD guidelines. We further present PoS tagging and dependency parsing
  experiments which report first results for the processing of the
  converted treebank. The full converted treebank was made available
  with the 1.2 release of the UD treebanks.\\
\newline
\Keywords{Dependency treebanks, Universal Dependencies, dependency parsing} }

\begin{document}
\maketitleabstract

\section{Introduction}
With the increasing popularity of dependency-based representations of
syntactic structure in recent years, a wealth of different dependency
annotation schemes have surfaced. It has been shown that the choice of
dependency scheme influences parsing results \cite{Sch:Abe:Rap:12} as
well as downstream applications \cite{Elm:Joh:Kle:13} and even though
attempts have been made to contrast different schemes theoretically
\cite{Iva:Oep:Ovr:12}, it is clear that the diversity of
representation makes comparisons difficult. Cross-linguistically even
more so, and it can often be difficult to tease apart
aspects of annotation scheme from typological differences in cross-lingual learning \cite{Soe:11,Skj:Ovr:12}.

Universal Dependencies (UD) \cite{Mar:Doz:Sil:14,Niv:15} is a recent
community-driven effort to create cross-linguistically consistent
syntactic annotation. UD is based on the Stanford dependency scheme
\cite{Mar:Mac:Man:06} which has become a widely used dependency scheme
for English in recent years.  A number of existing dependency
treebanks have been converted to UD \cite{Pyy:Kan:Miss:15,Niv:14}
and new data has also been annotated from scratch in order to enable
multilingual parser development, cross-lingual learning and
typological studies of syntactic structure. Treebanks involved in this
effort represent a diverse range of languages such as English, German,
Swedish, Spanish, Italian, Persian, Japanese, and the UD release 1.2
contains treebanks for as many as 33 different languages of varying
sizes.

This paper describes a fully automatic conversion procedure for the
Norwegian Dependency Treebank (NDT) to UD. Due to differences both in
the tag set, as well as structural analyses, the conversion requires
non-trivial transformations of the dependency trees, in addition to
mappings of tags and labels that make reference to a combination of
various kinds of linguistic information.  This paper details the
mapping of PoS tags, morphological features and dependency relations
and provides a description of the structural changes made to NDT
analyses in order to make it compliant with the UD guidelines. We
further present PoS-tagging and dependency parsing experiments which
contrast the performance of a state-of-the-art PoS-tagger and
dependency parser on the original and converted treebank. The full converted
treebank was made available with the 1.2 release of the UD treebanks
\cite{Niv:Mar:Gin:16}, which contains 37 treebanks for a total of 33 different languages.


\section{NDT and UD}
%138
The Norwegian Dependency Treebank (NDT) \cite{Sol:Skj:Ovr:14} contains
morphological and syntactic dependency annotation for both varieties
of written Norwegian (Bokm{\aa}l and Nynorsk).\footnote{The current
  Norwegian UD treebank contains the data from the Bokmål section of
  the treebank, which consists of 311,000 tokens. There are plans to
  include the Nynorsk data in the next UD release.} The morphological
annotation follows the Oslo-Bergen Tagger scheme
\cite{Hag:Joh:Nok:00}.  The syntactic annotation scheme is, to a large
extent, based on the Norwegian Reference Grammar
\cite{Faa:Lie:Van:97} and the dependency representations are
inspired by choices made in comparable treebanks, in particular the
Swedish treebank Talbanken \cite{Niv:Nil:Hal:2006}.
\newcite{Skj:14} quantified inter-annotator agreement using a chance-corrected metric derived from
Krippendorff's $\alpha$ and showed that agreement on the NDT data is high: scoring an $\alpha$ of about
98\%, among the highest of all the data sets studied.
The annotation guidelines \cite{Kin:Sol:Eri:2013} describe the annotation scheme in some detail and Table \ref{tb:choices} summarizes the main annotation choices in NDT \cite{Sol:Skj:Ovr:14}.

%45
\ctable[botcap,
    caption={Annotation choices in the NDT},
    label=tb:choices,
]{ll}{}{
        \FL
\multicolumn{1}{c@{\hspace{1em}}}{Head} & \multicolumn{1}{c@{\hspace{0.5em}}}{Dependent} \ML
Finite verb & Complementizer \NN
Finite auxiliary & Lexical verb \NN
Infinitival marker & Lexical verb \NN
Preposition & Prepositional complement \NN
Noun & Determiner\NN
First conjunct & Subsequent conjuncts
        \LL
    }

Universal Dependencies (UD) builds on several previous initiatives for
universally common morphological \cite{Zem:08,Pet:Das:McD:12} and
syntactic dependency \cite{McD:Niv:Qui:13,Ros:Mas:Mar:14}
annotation. Among its main tenets are the primacy of content-words,
i.e. content words, as opposed to function words, are syntactic heads
wherever possible. It is intended to be a universal annotation scheme,
i.e. applicable to any language, however also offers some
possibilities for language-specific information. With reference to the
NDT annotation choices in Table \ref{tb:choices}, the UD scheme adopts
the reverse attachment for auxiliaries, infinitival markers and
prepositions.

\section{Parts-of-speech}
% 78 + 287 = 365
The part-of-speech tag set used in the UD scheme is based on the
Universal PoS tag set of \newcite{Pet:Das:McD:12} and contains 17 tags. The NDT tag set
contains 19 tags.  The conversion of the part-of-speech information in
NDT to the UD PoS tag set is fairly straightforward and largely relies
on a direct mapping presented in Table \ref{tb:pos}.  A few
parts-of-speech require conversion rules which make reference to
additional information in the treebank, represented by disjunction
in the mapping. Below we will discuss a few of these cases.

%122
\ctable[botcap,
    caption={Mapping between NDT and UD parts-of-speech},
    label=tb:pos,
]{ll}{}{
        \FL
NDT & UD\ML
{\tt adj} & {\tt ADJ}\NN
{\tt adv} & {\tt ADV}\NN
{\tt clb} & {\tt PUNCT}, {\tt SYM}\NN
{\tt det} & {\tt DET}, {\tt NUM}\NN
{\tt konj} & {\tt CONJ}\NN
{\tt interj} & {\tt INTJ}\NN
{\tt inf-merke} & {\tt PART}\NN
{\tt prep} & {\tt ADP}, {\tt ADV}\NN
{\tt pron} & {\tt PRON}\NN
{\tt <komma>} & {\tt PUNCT}\NN
{\tt sbu} & {\tt SCONJ}\NN
{\tt <strek>} & {\tt PUNCT}\NN
{\tt subst} & {\tt NOUN}, {\tt PROPN}\NN
{\tt <anf>} & {\tt PUNCT}\NN
{\tt <parentes-slutt>} & {\tt PUNCT}\NN
{\tt <parentes-beg>} & {\tt PUNCT}\NN
{\tt symb} & {\tt SYM}\NN
{\tt ukjent} & {\tt X}\NN
{\tt verb} & {\tt AUX}, {\tt VERB}
        \LL
    }

The universal scheme makes a distinction between proper and common
nouns at the part-of-speech level. This information can be found among
the morphological features in NDT ({\tt prop}), hence the mapping is
straightforward.

For verbs, UD distinguishes auxiliaries ({\tt AUX}) from main
verbs ({\tt VERB}).  This distinction is not explicitly made in
NDT, hence our conversion procedure must make use of the syntactic
structure of the verbs in order to implement this distinction. Verbs
that have a direct, non-finite dependent (a dependent with the NDT
dependency relation {\sc inf}) are marked as auxiliaries and all other
verbs as regular verbs.

The relative pronoun {\it som} 'who/that' is analyzed as a subjunction
in NDT, whereas the universal scheme, and thus our conversion, uses
the pronominal tag {\tt PRON} for these words.

NDT explicitly marks headings using the {\tt |} symbol, tagged as a
clause boundary ({\tt clb}) along with information in the
morphological features ({\tt <overskrift>}). These are converted to the
UD symbol tag {\tt SYM}.

Numerical expressions, e.g. {\it to} 'two' or {\it 45} are tagged as
determiners in NDT, also when they do not explicitly modify a nominal,
e.g. {\it de to gjorde en imponerende innsats} 'the two did an
impressive job'. These are however marked as {\tt kvant}
(quantificational) in the morphological features, hence the conversion
mapping makes reference to this property in order to distinguish {\tt NUM} from {\tt DET}.

NDT implements a somewhat broader definition of prepositions than the UD
scheme and includes several cases which are counted as adverbs in the
universal schemes. In particular, this is the case for demonstrative
adverbs such as {\it her} 'here' and {\it der} 'there'. These are
followingly converted to adverbs ({\tt ADV}) in the PoS conversion, whereas all
other prepositions are converted to {\tt ADP}.

\section{Morphological information}
%109
In addition to part-of-speech information, NDT contains a rich
inventory of morphological features, e.g. information about properties
like gender, definiteness, tense, voice, etc. The UD guidelines
specify a universal set of morphological features and the conversion
between the two does not require reference to
information in addition to the feature information. The feature
mapping is described in Table \ref{tb:feats}.  Note that since the
number of UD features is larger than the NDT features, some of the NDT
features correspond to a set of UD features, e.g. the NDT features for
verbs ({\tt pres, pret}) which instantiate both the {\tt Mood, Tense}
and {\tt VerbForm} features.

%109
\ctable[botcap,
    caption={Mapping between NDT and UD morphological features},
    label=tb:feats,
]{lp{4cm}}{}{
        \FL
NDT & UD\ML
{\tt mask,fem,n{\o}yt} & {\tt Gender=Masc,Fem,Neut}\NN
{\tt ent,fl} & {\tt Number=Sing,Plur}\NN
{\tt be,ub} & {\tt Definite=Def,Ind}\NN
{\tt pres,pret} & {\tt Mood=Ind, Tense=Pres,Past, VerbForm=Fin}\NN
{\tt perf-part} & {\tt VerbForm=Part}\NN
{\tt imp} & {\tt Mood=Imp, VerbForm=Fin}\NN
{\tt pass} & {\tt Voice=Pass}\NN
{\tt inf} & {\tt VerbForm=Inf}\NN
{\tt 1,2,3} & {\tt Person=1,2,3}\NN
{\tt nom,akk,gen} & {\tt Case=Nom,Acc,Gen}\NN
{\tt pos,komp,sup} & {\tt Degree=Pos,Cmp,Sup}\NN
{\tt hum} & {\tt Animacy=Anim}\NN
{\tt pers} & {\tt PronType=Prs}\NN
{\tt dem} & {\tt PronType=Dem}\NN
{\tt sp} & {\tt PronType=Int}\NN
{\tt res} & {\tt PronType=Rcp}\NN
{\tt poss} & {\tt Poss=Yes}\NN
{\tt refl} & {\tt Refl=Yes}
        \LL
    }

\section{Structural conversion}
%63
The NDT annotation scheme differs structurally from the UD scheme in
a number of important ways.  The conversion is therefore non-trivial
and requires a set of structural rules which operate on the dependency
graphs in addition to a mapping procedure over the dependency labels.
The conversion is implemented as a cascade of structural rules followed by a
relation conversion procedure over the modified graph structures.
The structural rules employ a small set of graph operations that reverse, reattach, delete and add arcs.

\subsection{Root}
%52
In NDT, there is no designated root label. Rather, the root of the
dependency graph may have different labels, e.g. {\tt FINV} (finite
verb) or {\tt FRAG} (fragment), depending on the structure of the
sentence. In the conversion, every node with the dummy node (0) as
head receives the {\tt root} relation.

\subsection{Verbal groups}
%106
NDT consistently marks the finite verb as head of a clause, with other
non-finite verbs as dependents ({\tt INFV}), see example
(\ref{ex:verb}a). In a parallel manner, infinitival markers are also
annotated as heads with the infinitival verb as its dependent, see
(\ref{ex:infmark}a). UD on the other hand annotates the lexical, main
verb as head of the verbal group and various finite and non-finite
auxiliaries receive an auxiliary relation ({\tt aux, auxpass}) , see
(\ref{ex:verb}b) and (\ref{ex:infmark}b) below.  The conversion rule
locates the main verb within the chain of nonfinite dependents of the
finite verb and makes this node the head of the other verbs in the
chain.

%134
\begin{examples}
\item\label{ex:verb}
\begin{minipage}[t]{3cm}
a)
    \begin{dependency}
%    \begin{dependency}[arc edge, text only label, theme=simple]
%        \begin{deptext}
        \begin{deptext}[column sep=.2cm]
            skal \& ha \& sett \\
            \w{shall} \& \w{have} \& \w{seen}\\
        \end{deptext}
    \depedge{1}{2}{\sc infv}
    \depedge{2}{3}{\sc infv}
    \deproot[edge unit distance=1.5ex,hide label]{1}{}
    \end{dependency}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{3cm}
b)
%\item\label{ex:ndt-verb}
    \begin{dependency}
%    \begin{dependency}[arc edge, text only label, theme=simple]
        \begin{deptext}[column sep=.3cm]
            skal \& ha \& sett \\
            \w{shall} \& \w{have} \& \w{seen}\\
        \end{deptext}
    \depedge{3}{1}{aux}
    \depedge{3}{2}{aux}
    \deproot[edge unit distance=1.5ex,hide label]{3}{\large }
    \end{dependency}
\end{minipage}
\item\label{ex:infmark}
\begin{minipage}[t]{3cm}
a)
    \begin{dependency}
%    \begin{dependency}[arc edge, text only label, theme=simple]
        \begin{deptext}[column sep=.2cm]
            {\aa} \& ha \& sett \\
            \w{to} \& \w{have} \& \w{seen}\\
        \end{deptext}
    \depedge{1}{2}{\sc infv}
    \depedge{2}{3}{\sc infv}
    \deproot[edge unit distance=1.5ex,hide label]{1}{}
    \end{dependency}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{3cm}
b)
    \begin{dependency}
%    \begin{dependency}[arc edge, text only label, theme=simple]
        \begin{deptext}[column sep=.5cm]
            {\aa} \& ha \& sett \\
            \w{to} \& \w{have} \& \w{seen}\\
        \end{deptext}
    \depedge{3}{1}{mark}
    \depedge{3}{2}{aux}
    \deproot[edge unit distance=1.5ex,hide label]{3}{}
    \end{dependency}
\end{minipage}
\end{examples}


\subsection{Copula constructions}
%42
The treatment of copula constructions within the UD scheme differs
markedly from that of the NDT by appointing the predicative element as
head of the entire construction and attaching the copula verb with a
special relation {\tt cop}, see (\ref{ex:cop}b). Our conversion thus
reverses the arc between the copula and its complement and reattaches
all its dependents to the predicative element.
%68
\begin{examples}
\item\label{ex:cop}
\begin{minipage}[t]{3cm}
a)
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
            de \& er \& fattige \\
            \w{they} \& \w{are} \& \w{poor}\\
        \end{deptext}
    \depedge{2}{1}{\sc subj}
    \depedge{2}{3}{\sc spred}
    \deproot[edge unit distance=2ex]{2}{\sc finv}
    \end{dependency}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{3cm}
b)
%\item\label{ex:ndt-verb}
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.4cm]
            de \& er \& fattige\\
            \w{they} \& \w{are} \& \w{poor}\\
        \end{deptext}
    \depedge{3}{1}{nsubj}
    \depedge{3}{2}{cop}
    \deproot[edge unit distance=2ex]{3}{ root}
    \end{dependency}
\end{minipage}
\end{examples}


\subsection{Prepositions and their complements}
%47
In NDT, prepositions are heads of their prepositional complements
which receive the {\tt PUTFYLL} label, see (\ref{ex:prep}a). Seeing
that languages differ greatly in their use of pre/postpositions, the
UD scheme annotates the prepositional complement as head and attaches
the preposition using the {\tt case} relation, see (\ref{ex:prep}b). In the
conversion, this is once again a matter of reversing the arc and reattaching dependents to the new head.
%56
\begin{examples}
\item\label{ex:prep}
\begin{minipage}[t]{2.5cm}
a)
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
            p{\aa} \& menyen \\
            \w{on} \& \w{menu-the} \\
        \end{deptext}
    \depedge{1}{2}{\sc putfyll}
    \deproot[edge unit distance=1.5ex,hide label]{1}{}
    \end{dependency}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[t]{2.5cm}
b)
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
            p{\aa} \& menyen \\
            \w{on} \& \w{menu-the} \\
        \end{deptext}
    \depedge{2}{1}{case}
    \deproot[edge unit distance=1.5ex,hide label]{2}{}
    \end{dependency}
\end{minipage}
\end{examples}


\subsection{Predicatives}
%119
NDT distinguishes several types of predicatives, both predicatives
that are arguments of verbs (subject predicative {\tt SPRED} and
object predicative {\tt OPRED}) and ``free predicatives'' which are
not arguments of the verb, but nonetheless characterize either a
subject or an object in the preceding context (free subject
predicative {\tt FSPRED} and free object predicative {\tt
  FOPRED}). Both of these are attached to the finite verb in NDT, as we can see in (\ref{ex:pred}a). In a
similar manner, UD distinguishes between obligatory and optional
predicatives, where the former are analyzed using the {\tt xcomp}
relation and attached to the main predicate, whereas the optional
predicatives are attached as adverbial clauses ({\tt acl}) modifying the argument
they characterize, see (\ref{ex:pred}b). Our conversion thus attaches the {\tt FSPRED} argument to its sibling subject argument, {\tt FOPRED} to an object sibling.
%65
\begin{examples}
\item\label{ex:pred}
(a)\\
%\begin{itemize}
%\item[(a)]
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
          taler \& Solstad \& uredd \\
          \w{speaks} \& \w{Solstad} \& \w{unscared} \\
        \end{deptext}
            \depedge{1}{2}{\sc subj}
            \depedge{1}{3}{\sc fspred}
            \deproot[edge unit distance=1.5ex,hide label]{1}{}
    \end{dependency}
\\(b)\\
%\item[(b)]
%    \begin{dependency}[arc edge, text only label, theme=simple]
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
          taler \& Solstad \& uredd \\
          \w{speaks} \& \w{Solstad} \& \w{unscared} \\
        \end{deptext}
            \depedge{1}{2}{nsubj}
            \depedge{2}{3}{acl}
            \deproot[edge unit distance=1.5ex,hide label]{1}{}
    \end{dependency}
%\end{itemize}
\end{examples}


\subsection{Coordination}
%65
Coordination is a phenomenon which exhibits considerable variation in terms
of dependency representation across various annotation schemes \cite{Pop:Mar:Ste:13}.
As we can see in example in (\ref{ex:coord}), the analyses
in the NDT and UD schemes are fairly similar in their choice of
the first conjunct as head of the coordinate structure. They differ
mainly in the attachment of the conjunction and the relation names.
%88

\begin{example}
\label{ex:coord}
%\begin{itemize}
%\item[(a)]
 %   \begin{dependency}[arc edge, text only label, theme=simple]
(a)\\
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
          kamskjell \& , \& piggvar \& og \& lammefilet \\
          \w{scallops} \& , \& \w{turbot} \& \w{and} \& \w{lamb fillet}\\
        \end{deptext}
            \depedge{1}{2}{\sc ik}
            \depedge{1}{3}{\sc koord}
            \depedge[edge unit distance=2ex]{1}{5}{\sc koord}
            \depedge{5}{4}{\sc konj}
            \deproot[edge unit distance=2ex,hide label]{1}{}
    \end{dependency}
%\item[(b)]
%    \begin{dependency}[arc edge, text only label, theme=simple]
\\(b)\\
    \begin{dependency}
        \begin{deptext}[column sep=.2cm]
          kamskjell \& , \& piggvar \& og \& lammefilet \\
          \w{scallops} \& , \& \w{turbot} \& \w{and} \& \w{lamb fillet}\\
        \end{deptext}
            \depedge{1}{2}{punct}
            \depedge{1}{3}{conj}
            \depedge{1}{4}{cc}
            \depedge{1}{5}{conj}
            \deproot[hide label]{1}{}
    \end{dependency}
%\end{itemize}
%\item\label{ex:ndt-verb}
\end{example}

\section{Conversion of dependency relations}
%105
A minority of the dependency relations in NDT may be converted
directly, based on the mapping described in Table \ref{tb:dep}, the
rest require the formulation of mapping constraints which make
reference to information in addition to the dependency relation
itself, i.e. part-of-speech tag, morphological features, dependency
structure or even a combination of these. An overview of
the non-direct mapping of dependency relations that require additional
information from the linguistic context is provided in Table \ref{tb:multidep}.
These mappings often also require heuristics which approximate some
syntactic property which is not explicitly annotated in NDT. Below we
will present these heuristics and their usage in the conversion.

%102
\ctable[botcap,
    caption={Direct mapping between NDT and UD dependency relations.},
    label=tb:dep,
]{ll}{}{
        \FL
NDT & UD\ML
{\tt APP} & {\tt appos}\NN
{\tt FSUBJ} & {\tt expl}\NN
{\tt FOBJ} & {\tt expl}\NN
{\tt FSPRED} & {\tt acl}\NN
{\tt FOPRED} & {\tt acl}\NN
{\tt FRAG} & {\tt root}\NN
{\tt IOBJ} & {\tt iobj}\NN
{\tt OPRED} & {\tt xcomp}\NN
{\tt INTERJ} & {\tt discourse}\NN
{\tt KONJ} & {\tt cc}\NN
{\tt KOORD} & {\tt conj}\NN
{\tt KOORD-ELL} & {\tt remnant}\NN
{\tt IP} & {\tt punct}\NN
{\tt IK} & {\tt punct}\NN
{\tt PAR} & {\tt parataxis}\NN
{\tt SPRED} & {\tt xcomp}\NN
{\tt UKJENT} & {\tt goeswith}
        \LL
    }


%118
\paragraph{Active/Passive}There are two ways of expressing passive voice in Norwegian: a
morphological passive expressed by an addition of a {\it -s} to the
verb, e.g. {\it danses} 'to be danced' or a periphrastic passive which
is composed of the auxiliary {\it bli} 'to become' and a participle
form, e.g. {\it danset} 'danced'. Only the morphological passive is
marked explicitly as being in the passive voice.  We therefore define
a heuristic which counts a lexical main verb as passive, if it is (i)
a morphological passive, or (ii) is a participle headed by a form of
the auxiliary {\it bli}.  This heuristic is used in the conversion of
passive auxiliaries {\tt auxpass} and passive subjects {\tt nsubjpass,
  csubjpass}.

%128
\paragraph{Nominal/Clausal}Several of the UD relations assume a distinction between nominal and
clausal elements. This distinction is complicated somewhat by the fact
that in copula constructions, as described above, the complement of
the copula is head of the construction as a whole. This means that
adjectives or even nouns may be counted as clausal in contexts where
they have a copula dependent, as in (\ref{ex:cop}b).  In the
conversion we introduce a notion of a \emph{predicate}, which may be
either verbal ({\tt AUX, VERB}) or the complement in a copula
construction. This notion is used to distinguish nominal and clausal
subjects ({\tt nsubj, nsubjpass} vs. {\tt csubj, csubjpass}), objects
({\tt dobj} vs. {\tt ccomp, xcomp}), various modifiers ({\tt nmod}
vs. {\tt acl}) and adverbials ({\tt nmod} vs. {\tt advcl}).
%TO-DO: adverb subjects?? footnote

%59
\paragraph{Control}UD is inspired by the syntactic framework of Lexical Functional
Grammar \cite{Kap:Bre:82} and adopts its distinction between complement
clauses with obligatory subject control ({\tt xcomp}) and those without
({\tt ccomp}). The notion of control is not native to NDT, hence we
approximate it by requiring the presence of an explicit subject dependent of the
head verb of the clause.


%45
\paragraph{Negation}UD distinguishes negation modifiers which modify either a noun ({\it
  no problem}) or a predicate (in the aforementioned sense) ({\it is
  not a problem, doesn't argue}). Our conversion explicitly marks the
negative determiner ({\it ingen} 'no') and the negative adverb ({\it
  ikke} 'not') in Norwegian.

%141
\paragraph{Particles}NDT distinguishes between transitive and intransitive prepositions, or
so-called particles, in the annotation. In order to account for the
relation between the verb and its particle we introduce the
language-specific relation {\tt compound:prt}, for prepositions which
are attached to a verb and furthermore does not have an explicit prepositional complement,
e.g. {\it si opp} 'discontinue' (lit. 'say up').
On manual inspection of the
converted data, we find that this conversion also gives us the
preposition in fixed expressions which require the combination of a
verb, direct object and a preposition, such as {\it snu ryggen til}
'turn (one's) back to' as well as certain fixed combinations of
prepositions which function as a whole syntactically, e.g. {\it til og
  med} 'even (lit. to and with)'. Without any existing annotation,
distinguishing these from the regular particle constructions automatically is difficult.


\paragraph{Relative clauses}The analyses of relative clauses differ notably between the two
annotation schemes. Both schemes treat the relative clause as a
clausal modifier of a nominal element ({\tt ATR} vs {\tt acl}),
however, the treatment of the relative marker/pronoun differs. NDT
treats relative markers as subordinating conjunctions which depend on
the finite verb of the relative clause,\footnote{They argue for this
  by pointing to the fact that relative markers unlike many other
  languages do not conjugate (who-whom, der-die-das) and only occur
  initially in a subordinate clause \cite{Faa:Lie:Van:97}} while UD treats
the relative marker as a pronoun which occupies an argument relation
in the relative clause. In our conversion, we introduce a
language-specific variant of the clausal relation {\tt acl}, {\tt
  acl:relcl} which signals that this is a relative clause. The
relative marker {\it som} 'that' receives an argument role based on a
heuristic which checks for the presence or absence of other argument
siblings.

\ctable[botcap,
    caption={Non-direct mapping between NDT and UD dependency relations; requires additional constraints with reference to PoS, morphological features or dependency context.},
    label=tb:multidep,
]{lp{4cm}l}{}{
        \FL
NDT & UD \ML
{\tt ADV} & {\tt advcl, advmod, compound:prt, neg, nmod} \NN
{\tt ATR} & {\tt acl:relcl, amod, nmod} \NN
{\tt DET} & {\tt nmod, nummod, det} \NN
{\tt FINV} & {\tt aux, auxpass, root} \NN
{\tt FLAT} & {\tt foreign, name} \NN
{\tt OBJ, POBJ} & {\tt dobj, ccomp, xcomp} \NN
{\tt SBU} & {\tt nsubj, nsubjpass, dobj, iobj, mark} \NN
{\tt SUBJ, PSUBJ} & {\tt nsubj, nsubjpass, csubj, csubjpass}
        \LL
    }

\section{The converted treebank}
Following the conversion of the Norwegian Dependency Treebank to
Universal Dependencies scheme, 51.5\% of the tokens in the original treebank
were reattached. The resulting treebank contains 17 PoS tags and 35 different
morphological features for 311,277 tokens of Norwegian Bokmål.

All UD treebanks consist of three data sets: a training, development
and test set. In creating these data splits for Norwegian, care has
been taken to preserve contiguous texts in the different splits and
also to keep a balance of genres in each of the splits. Table
\ref{tb:splits} shows an overview of the Norwegian UD data sets in
terms of tokens and sentences.

\ctable[botcap,
    caption={Overview of the Norwegian UD train, dev and test data sets.},
    label=tb:splits,
]{lrr}{}{
        \FL
Data set & Tokens & Sentences\ML
no-ud-train & 244766 & 15696\\
no-ud-dev & 36467 & 2410\\
no-ud-test & 30034 & 1939\ML
Total & 20045 & 311277
        \LL
    }



\section{Tagging and parsing into UD}
Without a gold standard for Norwegian UD dependencies, it is difficult
to evaluate our conversion directly. We may however, evaluate
PoS-tagging and dependency parsing performance for the converted
treebank.  In the following, we report on a set of experiments which
investigate the performance of a state-of-the-art PoS-tagger and
parser trained and evaluated on the converted Norwegian UD
treebank. Even though the data sets are not strictly comparable, we
contrast the performance of the taggers and parsers with the results
for the original NDT scheme. We further investigate the effect of
using the automatically assigned PoS-tags during parsing in order to
achieve a maximally realistic setting.

\subsection{PoS-tagging}
For PoS-tagging, we experiment with two different tag sets: a coarse
tag set, consisting of the simple PoS-tags in the two schemes (NDT and
UD), and a fine-grained set, which represents the concatenation of
the PoS-tag and the morphological features. The original tag set of NDT
comprises 19 tags, 12 of which are morphosyntactic tags\footnote{The
  remaining tags describe various types of punctuation, such as
  commas, dashes, etc.}, while its fine-grained counterpart totals 368
tags. For UD, the coarse-grained tag set consists of 17 tags and the
corresponding fine-grained tag set of UD comprises 169 tags.

The PoS-tagging was performed by
SVMTool\footnote{\url{cs.upc.edu/~nlp/SVMTool}} \cite{Gim:Mar:04},
employing strategy 1, which proved to be optimal for tagging NDT in
previous work \cite{Hohle:16}. All experiments are performed on the
data splits detailed in Table \ref{tb:splits}.

The results for PoS-tagging are presented in Table
\ref{tb:taggingeval}. The PoS-tagger obtains an accuracy of 96.96\% on
the dev set and 96.82\% on the test set. For the fine-grained UD set,
the performance drops to 94.59\% (dev) and 94.15\% (test).

In general, we observe that tagging accuracy is higher on NDT with the
original tag set, while UD surpasses NDT when using the fine-grained
tag set. This is as expected, as the fine-grained tag set of NDT
contains almost 200 more tags than that of UD, which markedly
complicates the tagging.

An error analysis reveals that for both schemes, the recognition of
interjections ({\tt interj, INTJ}) and unknown/foreign words ({\tt
  ukjent, X}) are difficult. We further note that the UD distinction
between auxiliaries and main verbs seems difficult to make. For the
NDT-tagger, the {\tt verb} tag has an F-score of 97\%, whereas the UD
{\tt VERB} and {\tt AUX} tags receive F-scores of 94\% and 91\%,
respectively.

\ctable[botcap,
    caption={Overview of the results from tagging NDT and UD with the various
        data sets and tag set.},
    label=tb:taggingeval,
]{llll}{}{
    \FL
    & & \multicolumn{2}{c}{Accuracy}
    \NN
    \cmidrule(lr){3-4}
    Data set & Tag set & NDT & UD
    \ML
    Dev & Coarse & 97.90\% & 96.96\% \\
    Dev & Fine & 93.74\% & 94.59\% \\
    Test & Coarse & 97.82\% & 96.82\% \\
    Test & Fine & 93.19\% & 94.15\%
    \LL
}

\subsection{Dependency parsing}
We perform a set of parse experiments in order to evaluate the UD
scheme for Norwegian, and further compare these results to those
obtained using the original NDT schemes. In our experiments, we
further contrast the use of gold vs. automatically assigned PoS-tags
during training and testing. For the experiments using automatically
assigned PoS-tags, the parser was both trained and tested on
automatically assigned tags, as this proved to give slightly better
results on the development set.

All experiments are performed on the data splits detailed in Table
\ref{tb:splits}. For the experiments, we employ the Mate
parser\footnote{\url{code.google.com/archive/p/mate-tools}}
\cite{Boh:10}, as it proved best in previous work on parsing of NDT
\cite{Sol:Skj:Ovr:14}. It was run on dev and test sets in separate
experiments, with either gold or automatic tags (i.e., the
coarse-grained tags as predicted by SVMTool as described above).  When
using automatic tags, the morphological features are excluded from the
data set, for the most realistic comparison.

The parsing results are presented in Table \ref{tb:parsingeval}. We
find that the UD-parser obtains a labeled accuracy score (LAS) of
88.5\% on both the dev and test sets using gold standard PoS-tags. The
corresponding results for the NDT-parser are 90.15\% (dev) and 90.55\%
(test).

\ctable[botcap,
    caption={Overview of the results from parsing NDT and UD with the various
        data sets and tag setups.},
    label=tb:parsingeval,
]{llllll}{}{
    \FL
    & & \multicolumn{2}{c}{LAS} & \multicolumn{2}{c}{UAS}
    \NN
    \cmidrule(lr){3-4}
    \cmidrule(lr){5-6}
    Data & Tags & NDT & UD & NDT & UD
    \ML
    Dev & Gold & 90.15\% & 88.50\% & 92.51\% & 91.13\% \\
    Dev & Auto & 86.73\% & 83.91\% & 89.99\% & 87.16\% \\
    Test & Gold & 90.55\% & 88.54\% & 92.76\% & 91.21\% \\
    Test & Auto & 86.76\% & 83.86\% & 90.13\% & 87.16\%
    \LL
}

The drop in results compared to the NDT scheme is not surprising. As
noted in previous work \cite{Mar:Doz:Sil:14}, several of the design
choices of the UD scheme, such as the attachment of auxiliaries and
prepositions to content words, are known to be more difficult to parse
than their reverse counterparts \cite{Sch:Abe:Rap:12}. The UD scheme
is not primarily designed to give optimal parse results and is further
designed to be cross-lingually applicable. A similar trend was
observed for Danish UD parsing \cite{Joh:Alo:Pla:15}. They observed a
drop in results from the original 84.38\% LAS to 81.56 \% LAS for UD.
For Bulgarian, \newcite{Ose:Sim:15} observed a drop from 89.14\% to
83.5\% for a preliminary version of their UD conversion.

In the experiments using automatically assigned tags (Auto), we
combine the output of the taggers described in the previous section
with the NDT and UD parsers. As expected, we observe a drop in parse
results when we switch to automatically assigned PoS-tags. For the
UD-parser, we observe a drop of 4.6 percentage points, whereas the
corresponding figure for the NDT-parser is 3.4 percentage points (on
the dev sets). It is clear that the somewhat lower tagging accuracy
for the UD treebank influences parsing results.

An error analysis in terms of dependency relation assignment and
attachment, shows that the distinction of various clausal dependency
relations ({\tt acl, advcl, csubj, xcomp, ccomp}) are challenging for
the parser. Also, the assignment of the expletive relation {\tt expl}
is challenging for the UD parser. Since the neuter pronoun {\it det}
'it' may be both referential and non-referential in Norwegian and has the same
PoS-tag regardless of referentiality, this clearly constitutes a
complex distinction.

\section{Conclusion}
This article has presented a fully automatic conversion of the
Norwegian Dependency Treebank to Universal Dependencies. The
conversion consists of a set of mappings for PoS tags, morphological
features and dependency relations as well as structural rewrite rules
that transform the NDT analyses to Universal Dependencies. In order to
evaluate the conversion we employ the treebank in a set of tagging and
parsing experiments which show that even though the results are lower
than for the original scheme, the UD version is still as viable option
for processing of Norwegian text. The results are shown to be in line
with those observed for several other UD converted treebanks.

This article has presented the first release of the Norwegian UD
treebank. The conversion is still under development, and we plan to
release new versions along with the coming releases of the UD
treebank. An important line of work in the time to come is to increase
the consistency of the UD treebanks, and efforts are being made to
make the Norwegian data set more consistent with the other Germanic
treebanks in particular. We also plan to include the data for the
other written variant of Norwegian (Nynorsk) from the original NDT
treebank in one of the next releases.

\section{References}
\label{main:ref}

\bibliographystyle{lrec2016}
\bibliography{lrec2016}


%% \section{Language Resource References}
%% \label{lr:ref}
%% \bibliographystylelanguageresource{lrec2016}
%% \bibliographylanguageresource{lrec2016}

%\clearpage
%\smaller
%\bibliographystyle{lrec2016}
%\bibliography{lrec2016}

\end{document}





